---
title: "Homework 6"
author: "Denis Ostroushko"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    extra_dependencies: ["float"]
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo = F}
knitr::opts_chunk$set(echo = T, message = F, warning = F, fig.pos = "!H")
options(scipen=999)
```

```{r, include=F}
library(MASS) # for stepwise regression
  # pro tip: load MASS before tidyverse, otherwise it really messes with select() and other essential data wrangling 
  # function 
library(tidyverse)
library(kableExtra)
library(readxl)
library(gridExtra)
library(ggeffects)
library(mltools) # one hot encoding outside of caret package 
library(data.table) # need this for mltools to work 
library(olsrr) #another package for stepwise regression 
```

```{r, eval=F}
library(MASS) # for stepwise regression
  # pro tip: load MASS before tidyverse, otherwise it really messes with select() and other essential data wrangling 
  # function 
library(tidyverse)
library(kableExtra)
library(readxl)
library(gridExtra)
library(ggeffects)
library(mltools) # one hot encoding outside of caret package 
library(data.table) # need this for mltools to work 
library(olsrr) #another package for stepwise regression 
```


# 12.2 

```{r}
infants <- readxl::read_xls('/Users/denisostroushko/Desktop/UofM MS/MS Fall 2022/Puhb 7405/Data Sets/Infants.xls')

colnames(infants) <- c("head_c", "length", "gest_weeks", "birth_w", "m_age", "toxemia")

# process the data and keep variables for analysis 

infants_f <- infants %>% 
  select(birth_w, gest_weeks, m_age)

```

## 12.2 - A

### Model Specifications and T-tests

```{r, echo = F}
ggplot(data = infants_f, 
       aes(x = m_age, 
           y = birth_w)) + geom_point() + 
  
  stat_smooth(se = F, aes(color = "Smooth Trend Line")) + 
  stat_smooth(se = F, method = "lm", aes(color = "Regression Line")) + 
  
  scale_color_manual(values = c("Smooth Trend Line" = "blue", 
                                "Regression Line" = "red")) + 
  xlab("Mother's Age") + 
  ylab("Infant's Birth Weight") + 
  ggtitle(paste("Correlation Between Mother's Age  \n and Infant's Birth Weight: ", 
                round(cor(infants_f$m_age, 
                          infants_f$birth_w), 4))) + 
  theme_minimal()

```

```{r, echo = F}
# look at the correlation between age and age^2 
ggplot(data = infants_f, 
       aes(x = m_age, 
           y = m_age^2 )) + geom_point() + 
  
  stat_smooth(method = "lm", se = F, aes(color = "Fitted Regression Line")) + 
  scale_color_manual(values = c("Fitted Regression Line" = "red")) + 
  
  xlab("Age") + 
  ylab("Age Squared") + 
  ggtitle(paste("Correlation: ", round(cor(infants_f$m_age, infants_f$m_age^2),4))) + 
  theme_minimal()

```

Model specification: 

$$\Large E[Y] = \hat \beta_0 + \hat \beta_1 * Gestional \ Weeks + \hat \beta_2 * Mother's \ Age + \hat \beta_3 * Mother's \ Age ^ 2$$
Model Summary 

```{r, echo = F}
inf_lm <- lm(birth_w ~ gest_weeks + m_age + I(m_age^2), data = infants_f)

sum_data <- data.frame(summary(inf_lm)$coefficients)

sum_data$names <- c("Intercept", "Gestational Weeks", "Mother's Age", "Mother's Age Squared")

rownames(sum_data) <- NULL

sum_data <- sum_data %>%  dplyr::select(names, everything())

round_3 <- function(x){round(x,3)}
sum_data[,2:5] <- lapply(sum_data[,2:5], round_3)

colnames(sum_data) <-c("Model Term", "Estimate", "Std. Error", "T-value", "P-value") 

sum_data %>% 
  kbl(booktabs = T, align = c('l', 'c','c','c','c')) %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))

```

Comemnts on Model summary: 

* R and Adjusted R: `r round(summary(inf_lm)$r.squared,4)` 
`r round(summary(inf_lm)$adj.r.squared,4) `

* Coefficients for Age and Age ^ 2

### Evaluate Extra Sum of Squares

Focus: Evaluate SSR(Age^2 | Gest, Age)

```{r, echo = F}
anv <- data.frame(anova(inf_lm) )

anv$names <- c("Gestational Weeks", "Mother's Age", "Mother's Age Squared", "Residuals")
rownames(anv) <- NULL
anv <- anv %>% select(names, everything())

colnames(anv) <- c("Model Term", "DF", "SS", "MS", "F-statistic", "P(F* > F)")

anv[,length(anv)] <- round(anv[,length(anv)], 4)

anv %>% 
  kbl(booktabs = T, align = c('l', rep('c', length(anv)-1)) ) %>% 
  kable_styling(latex_options = c("striped", "HOLD_position")) %>% 
  row_spec(nrow(anv)-1, hline_after = T)

```

```{r, echo = F}
# R^2 (m age | gest weeks )
lm_2 <- lm(birth_w ~ gest_weeks + m_age, data = infants_f)
ssr_2 <- sum( (lm_2$fitted.values - mean(infants_f$birth_w) )^2 )

# R^2 (m age ^2  | m age, gest weeks )
lm_3 <- lm(birth_w ~ gest_weeks + m_age + I(m_age^2), data = infants_f)

sse_2 <- sum(lm_2$residuals^2)
ssr_3 <- sum( (lm_3$fitted.values - mean(infants_f$birth_w) )^2 )

Rs_3 <- (ssr_3 - ssr_2) /sse_2

```

* Extra SS

* Extra R^2 

* Connection with the t-test 

### Visualize Model Effects

```{r, echo = F }
mydf <- ggpredict(inf_lm, terms = "m_age")

mean_name <- paste0("Birth Weight Mean Value: ", round(mean(infants_f$birth_w)))

ggplot(mydf, aes(x, predicted)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = .1) + 
  theme_minimal() + 
  geom_hline( 
             aes(yintercept = mean(infants_f$birth_w),color = mean_name)
             ) + 
  guides(color=guide_legend(title="Additional Elements: ")) + 
  theme(
    legend.position = "bottom"
  ) + 
  
  ylab("Predicted Values of Birth Weight") + 
  xlab("Mother's Age") + 
  ggtitle("Model Estiamted Effects of Mother's Age on \n Infant's Birth Weight")
  

```

* Comment on Standard Error and fit, we can fit a line with slope =

### Interpretation of Mother's Age Coefficients 

From google, interpretation of the quadratic coefficient: 

"
A positive quadratic coefficient causes the ends of the parabola to point upward. A negative quadratic coefficient causes the ends of the parabola to point downward. The greater the quadratic coefficient, the narrower the parabola. The lesser the quadratic coefficient, the wider the parabola."

https://stats.stackexchange.com/questions/108657/how-to-interpret-coefficients-of-x-and-x2-in-same-regression

It may be useful to describe the effect of a unit change at some low value, some high value and somewhere in between.

## 12.2 - B

Correlation Transformation for variables $Y, X_1, ..., X_{p-1}$, denoted by $V$: 

$$\Large V^* = \frac{1}{\sqrt{n - 1}} \times \Big (\frac{V - \bar V}{sd(V)} \Big )$$
```{r}

correlation_transformation <- 
  function(X, n = nrow(infants_f_cor_tr)){
    
    1/(sqrt(n - 1)) * (X - mean(X))/sd(X)
    
  }

infants_f$m_age_sq <- infants_f$m_age^2
infants_f_cor_tr <- infants_f

infants_f_cor_tr <- data.frame(lapply(infants_f_cor_tr, correlation_transformation))
```

```{r, echo = F}
cor_tr_lm <- lm(birth_w ~ . , data = infants_f_cor_tr)
sum2 <- data.frame(summary(cor_tr_lm)$coefficients)

sum2$names <- c("Intercept", "Gestational Weeks", "Mother's Age", "Mother's Age Squared")

rownames(sum2) <- NULL

sum2 <- sum2 %>%  dplyr::select(names, everything())

round_3 <- function(x){round(x,3)}
sum2[,2:5] <- lapply(sum2[,2:5], round_3)

colnames(sum2) <-c("Model Term", "Estimate", "Std. Error", "T-value", "P-value") 

kbl(sum_data, booktabs = T, caption = "Original Scale Regression Estimates") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))

kbl(sum2, booktabs = T, caption = "Correlation  Transformation Regression Estimates") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))

```

* intercept is zero as expected in corr transformed 

* P-values are different for m age 

* Same conclusions apply 

## 12.2 - C

Transformation back to the original scale: 

For variables $X_1, ..., X_{p-1}$: 

$$\Large \hat \beta_i =  \hat \beta^*_i \times \frac{sd(Y)}{sd(X_i)}$$
```{r}
transform_back <-
  function(Beta_star, s_x, s_y){
    Beta_star * (s_y / s_x)
  }

S_Y <- sd(infants_f$birth_w)
```

Hid code to prepare the table. 

```{r, echo = F}
check_res <- 
  data.frame(
    names = c("Gestation Weeks", "Mother's Age", "Mother's Age Squared"), 
    betas = c(transform_back(sum2[2,2], s_x = sd(infants_f$gest_weeks), s_y = S_Y),
              transform_back(sum2[3,2], s_x = sd(infants_f$m_age), s_y = S_Y),
              transform_back(sum2[4,2], s_x = sd(infants_f$m_age_sq), s_y = S_Y)
              ), 
    ci_l = c(transform_back(confint(cor_tr_lm)[2,1], s_x = sd(infants_f$gest_weeks), s_y = S_Y),
              transform_back(confint(cor_tr_lm)[3,1], s_x = sd(infants_f$m_age), s_y = S_Y),
              transform_back(confint(cor_tr_lm)[4,1], s_x = sd(infants_f$m_age_sq), s_y = S_Y)
              ),
    ci_h = c(transform_back(confint(cor_tr_lm)[2,2], s_x = sd(infants_f$gest_weeks), s_y = S_Y),
              transform_back(confint(cor_tr_lm)[3,2], s_x = sd(infants_f$m_age), s_y = S_Y),
              transform_back(confint(cor_tr_lm)[4,2], s_x = sd(infants_f$m_age_sq), s_y = S_Y)
              )
  )

```


recall the the original model with the transformed variables was called `inf_lm`. Used it for Extra SS,
t-tests and model effects. We can obtain standard errors and confidence intervals for the estimates
to compare with the transformation back from the correlation transformation procedure. 

```{r}
conf <- data.frame(confint(inf_lm)) # just the confidence intervals 
conf <- cbind(coefficients(inf_lm), conf )
```

```{r, echo=F}
conf$names <- c("Intercept", "Gestation Weeks", "Mother's Age", "Mother's Age Squared")
rownames(conf) <- NULL

conf <- conf %>% select(names, everything())

colnames(conf) <- c("Model Term", "Coefficient", "95% C.I. Lower Bound", "95% C.I. Upper Bound")

round_3 <- function(x){round(x,3)}
conf[,2:4] <- sapply(conf[, 2:4], round_3)

conf %>% filter(`Model Term` != "Intercept" ) %>% 
  kbl(align = c('l', rep('c', length(conf)-1)), booktabs =  T, 
      caption = "Original Model Estiamtes and C.I.") %>% 
  kable_styling(latex_options = c("HOLD_posiiton", "striped")) 

```

```{r, echo = F}

colnames(check_res) <- c("Model Term", "Coefficient", "95% C.I. Lower Bound", "95% C.I. Upper Bound")

check_res[,2:4] <- lapply(check_res[,2:4], round_3)

check_res %>% filter(`Model Term` != "Intercept" ) %>% 
  kbl(align = c('l', rep('c', length(conf)-1)), booktabs =  T, 
      caption = "Estimaes obtained via \n Back-Trnasformation and C.I.") %>% 
  kable_styling(latex_options = c("HOLD_posiiton", "striped")) 
```

so we can use linear transformations 
good to know 

# 13.4 

```{r}
cig <- read_xlsx('/Users/denisostroushko/Desktop/UofM MS/MS Fall 2022/Puhb 7405/Data Sets/E-CID-3.xlsx')

cig$Y1 <- with(cig, log(NNAL_vt4_creat / NNAL_vt0_creat))
cig$Y2 <- with(cig, log(TNE_vt4_creat / TNE_vt0_creat))

cig <- cig %>% 
  select(Y1, Y2, arm, age, gender, white, educ2, income30, FTND)

colnames(cig)[length(cig)] <- "ftnd"
```

## 13.4 - A

* Arm will result in 4 -1 variables 

* Age is untouched 

* FTND is treated as continuous 

* Others need to be converted to factor variables 

```{r}

cig <- cig %>% select(
  Y1, Y2, age, arm, gender, educ2, income30, ftnd
)

cig$arm <- as.factor(cig$arm)

cig <- data.frame(one_hot(as.data.table(cig))) %>% select(-arm_5)

cig[,4:(length(cig)-1)] <- lapply(cig[,4:(length(cig)-1)], as.factor)
```

```{r}
n_unique <- function(x){length(unique(x))}

meta_data <- 
  
  data.frame(
    class = sapply(cig, class), 
    n_unique = sapply(cig, n_unique)
  )
```

```{r, echo = F}

meta_data$name <- rownames(meta_data)

rownames(meta_data) <- NULL

meta_data <- meta_data %>% filter(!(name %in% c("Y1", "Y2")))

meta_data <- 
  meta_data %>% select(name, everything())
  
colnames(meta_data) <- c("Predictors", "Assigned Class", "N of Unique Values")

meta_data %>% 
  kbl(caption = "Sumamry of Covariates",
       booktabs = T, align = c("l", 'c', 'c')) %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))

n_predictors <- nrow(meta_data)
```

## 13.4 - B

### Regression on Y1 

 add model specification 
 
```{r}
y1_lm1 <- lm(Y1 ~ . - Y2, data = cig )
```

```{r , echo = F}
sum2 <- data.frame(summary(y1_lm1)$coefficients)

sum2$names <- c("Intercept", "Age", "Arm 6", "Arm 7", "Arm 8", 
                "Gender", "Education", "Income >= $30K", "FTND")

rownames(sum2) <- NULL

sum2 <- sum2 %>%  dplyr::select(names, everything())

round_3 <- function(x){round(x,3)}
sum2[,2:5] <- lapply(sum2[,2:5], round_3)

colnames(sum2) <-c("Model Term", "Estimate", "Std. Error", "T-value", "P-value") 

kbl(sum2, booktabs = T, caption = "Original Scale Regression Estimates") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))
```

* Bonferroni Adjustments 

    + $P-value = 0.05$
    
    + $Bonferroni \ adjusted \ P-value =$ `r round(0.05 / n_predictors, 4)`

corrected p-value = p-value / number of predictors 

```{r}
sum_bonf_adj <- sum2 %>% select(`Model Term`, `P-value`)
sum_bonf_adj$`Significant at Adj. Level` = 
  with(sum_bonf_adj, 
       ifelse(`P-value` < 0.05 / n_predictors , "*", "")
       )

sum_bonf_adj %>% 
  kbl( booktabs = T, caption = "Regression of Y1 Bonferroni Adjusted Comparison") %>% 
    kable_styling(latex_options = c("striped", "HOLD_position")) %>% 
    column_spec(3, width = "2cm")

```

* HOLM Adjustments 

    + order p-values smallest to largest 
    + if first p-value if smaller than $0.05 / 8 =$ `r round(0.05 / n_predictors, 4)` then 
      conclude significance, and move to next predictor, otherwise stop, none are significant 
      
    + next predictor will be tested at $0.05 / 7 =$ `r round(0.05 / (n_predictors-1), 4)`

```{r}

holm_data <- 
  sum2 %>% select(`Model Term`, `P-value`) %>% arrange(`P-value`) %>% 
  filter(`Model Term` != "Intercept")
  
holm_data$`Comparison P-value` <- 1
holm_data$`Significant at Adj. Level` <- ""

cur_adj_n <- n_predictors

for(i in 1:nrow(holm_data)){
  
  cur_level <- 0.05 / cur_adj_n
  holm_data[i,3] <- cur_level
  
  if(holm_data[i,2] <= cur_level ){
    cur_adj_n <- cur_adj_n - 1
    holm_data[i,3] <- cur_level
    holm_data[i,4] <- "*"
  }
}

holm_data[,2:3] <- lapply(holm_data[,2:3], round_3)

holm_data %>% 
  kbl( booktabs = T, caption = "Regression of Y1 HOLM Adjusted Comparison") %>% 
    kable_styling(latex_options = c("striped", "HOLD_position")) %>% 
    column_spec(c(3,4), width = "2cm")

```

* Hochberg Adjustments

    + Sort P-values largest to smallest 
    + Compare the largest to $0.05$, if significant, declare all significant
    + Otherwise, compare the next one to $0.05/2 = 0.025$
    + Keep comparing to $0.05/3$, $0.05/4$, etc.. until we find a comparison where 

    
```{r}

hoch_data <- 
  sum2 %>% select(`Model Term`, `P-value`) %>% arrange(-`P-value`) %>% 
  filter(`Model Term` != "Intercept")

hoch_data$`Comparison P-value` <- 0.05
hoch_data$`Significant at Adj. Level` <- ""

cur_adj_n <- 1

for(i in 1:nrow(hoch_data)){
  
  cur_level <- 0.05 / cur_adj_n
  hoch_data[i,3] <- cur_level
 
  if(hoch_data[i,2] > cur_level){
    cur_adj_n <- cur_adj_n + 1
    
    holm_data[i,3] <- cur_level
  }
}

hoch_data[,4] <- ifelse(hoch_data[,2] < hoch_data[,3], "*", "")

hoch_data[,2:3] <- lapply(hoch_data[,2:3], round_3)

hoch_data %>% 
  kbl( booktabs = T, caption = "Regression of Y1 HOCHBERG Adjusted Comparison") %>% 
    kable_styling(latex_options = c("striped", "HOLD_position")) %>% 
    column_spec(c(3,4),  width = "2cm")

```

* SUMMARY OF COEFFICIENT SELECTION FOR Y1 REGRESSION 

### Regression on Y2 

```{r}
y2_lm1 <- lm(Y2 ~ . - Y1, data = cig )
```

```{r , echo = F}
sum2 <- data.frame(summary(y2_lm1)$coefficients)

sum2$names <- c("Intercept", "Age", "Arm 6", "Arm 7", "Arm 8", 
                "Gender", "Education", "Income >= $30K", "FTND")

rownames(sum2) <- NULL

sum2 <- sum2 %>%  dplyr::select(names, everything())

round_3 <- function(x){round(x,3)}
sum2[,2:5] <- lapply(sum2[,2:5], round_3)

colnames(sum2) <-c("Model Term", "Estimate", "Std. Error", "T-value", "P-value") 

kbl(sum2, booktabs = T, caption = "Original Scale Regression Estimates") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))
```

* Bonferroni Adjustments 

```{r}
sum_bonf_adj <- sum2 %>% select(`Model Term`, `P-value`)
sum_bonf_adj$`Significant at Adj. Level` = 
  with(sum_bonf_adj, 
       ifelse(`P-value` < 0.05 / n_predictors , "*", "")
       )

sum_bonf_adj %>% 
  kbl( booktabs = T, caption = "Regression of Y2 Bonferroni Adjusted Comparison") %>% 
    kable_styling(latex_options = c("striped", "HOLD_position")) %>% 
    column_spec(3, width = "2cm")

```

* HOLM Adjustments 

```{r}

holm_data <- 
  sum2 %>% select(`Model Term`, `P-value`) %>% arrange(`P-value`) %>% 
  filter(`Model Term` != "Intercept")
  
holm_data$`Comparison P-value` <- 1
holm_data$`Significant at Adj. Level` <- ""

cur_adj_n <- n_predictors

for(i in 1:nrow(holm_data)){
  
  cur_level <- 0.05 / cur_adj_n
  holm_data[i,3] <- cur_level
  
  if(holm_data[i,2] <= cur_level ){
    cur_adj_n <- cur_adj_n - 1
    holm_data[i,3] <- cur_level
    holm_data[i,4] <- "*"
  }
}

holm_data[,2:3] <- lapply(holm_data[,2:3], round_3)

holm_data %>% 
  kbl( booktabs = T, caption = "Regression of Y2 HOLM Adjusted Comparison") %>% 
    kable_styling(latex_options = c("striped", "HOLD_position")) %>% 
    column_spec(c(3,4), width = "2cm")

```

* Hochberg Adjustments
    
```{r}

hoch_data <- 
  sum2 %>% select(`Model Term`, `P-value`) %>% arrange(-`P-value`) %>% 
  filter(`Model Term` != "Intercept")

hoch_data$`Comparison P-value` <- 0.05
hoch_data$`Significant at Adj. Level` <- ""

cur_adj_n <- 1

for(i in 1:nrow(hoch_data)){
  
  cur_level <- 0.05 / cur_adj_n
  hoch_data[i,3] <- cur_level
 
  if(hoch_data[i,2] > cur_level){
    cur_adj_n <- cur_adj_n + 1
    
    holm_data[i,3] <- cur_level
  }
}

hoch_data[,4] <- ifelse(hoch_data[,2] < hoch_data[,3], "*", "")

hoch_data[,2:3] <- lapply(hoch_data[,2:3], round_3)

hoch_data %>% 
  kbl( booktabs = T, caption = "Regression of Y2 HOCHBERG Adjusted Comparison") %>% 
    kable_styling(latex_options = c("striped", "HOLD_position")) %>% 
    column_spec(c(3,4),  width = "2cm")

```

## 13.4 - C

### Step Wise Regression on Y1 

comment that we have 2^8 possible models, and these are candidates for the best possible model 

```{r}
k <- ols_step_best_subset(y1_lm1)

k %>% dplyr::select(n, predictors) %>% 
  kbl(booktabs = T, 
      caption = "Regression of Y1, Best Candidate Models") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))
```

```{r, echo = F}
g1 <- 
  ggplot(data = k, 
         aes(x = n, y = adjr)) + geom_point(size = 1) + 
    geom_line(color = "blue") + 
    theme_minimal() + 
  xlab("Model Index") + 
  ylab("Adjusted R-square") +
  ggtitle("Regression of Y1, \n Adjusted R-square for \n Candidate Models")

g2 <- 
  ggplot(data = k, 
         aes(x = n, y = aic)) + geom_point(size = 1) + 
    geom_line(color = "blue")+ 
    theme_minimal() + 
  xlab("Model Index") + 
  ylab("AIC") +
  ggtitle("Regression of Y1, \n AIC for Candidate Models")

grid.arrange(g1, g2, nrow = 1)
```

```{r, echo = F}

k <- data.frame(k)

k_f <- 
  k %>% filter(n == 3) %>% 
  dplyr::select(predictors, rsquare, adjr, aic)

colnames(k_f) <- c("Predictors", "R-squared", "Adj. R-squared", "AIC")

k_f[,2:4] <- lapply(k_f[,2:4], round_3)

k_f %>% 
  kbl(caption = "Regression of Y1, Parameters of Selected Model", 
      booktabs = T)

```

```{r, echo = F}
y1_lm1_step <- stepAIC(y1_lm1, direction = "both", 
                      trace = FALSE)
```


```{r, echo = F}
sum2 <- data.frame(summary(y1_lm1_step)$coefficients)

sum2$names <- c("Intercept", "Arm 6",  "Arm 8","Income >= $30K")

rownames(sum2) <- NULL

sum2 <- sum2 %>%  dplyr::select(names, everything())

round_3 <- function(x){round(x,3)}
sum2[,2:5] <- lapply(sum2[,2:5], round_3)

colnames(sum2) <-c("Model Term", "Estimate", "Std. Error", "T-value", "P-value") 

kbl(sum2, booktabs = T, caption = "") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))

```

### Step Wise Regression on Y2


```{r}
k <- ols_step_best_subset(y2_lm1)

k %>% dplyr::select(n, predictors) %>% 
  kbl(booktabs = T, 
      caption = "Regression of Y2, Best Candidate Models") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))

```

```{r, echo = F}
g1 <- 
  ggplot(data = k, 
         aes(x = n, y = adjr)) + geom_point(size = 1) + 
    geom_line(color = "blue") + 
    theme_minimal() + 
  xlab("Model Index") + 
  ylab("Adjusted R-square") +
  ggtitle("Regression of Y2, \n Adjusted R-square for \n Candidate Models")

g2 <- 
  ggplot(data = k, 
         aes(x = n, y = aic)) + geom_point(size = 1) + 
    geom_line(color = "blue")+ 
    theme_minimal() + 
  xlab("Model Index") + 
  ylab("AIC") +
  ggtitle("Regression of Y2, \n AIC for Candidate Models")

grid.arrange(g1, g2, nrow = 1)
```

```{r, echo = F}

k <- data.frame(k)

k_f <- 
  k %>% filter(n == 2) %>% 
  dplyr::select(predictors, rsquare, adjr, aic)

colnames(k_f) <- c("Predictors", "R-squared", "Adj. R-squared", "AIC")

k_f[,2:4] <- lapply(k_f[,2:4], round_3)

k_f %>% 
  kbl(caption = "Regression of Y1, Parameters of Selected Model", 
      booktabs = T)

```

```{r, echo = F}
y2_lm1_step <- stepAIC(y2_lm1, direction = "both", 
                      trace = FALSE)

sum2 <- data.frame(summary(y2_lm1_step)$coefficients)

sum2$names <- c("Intercept", "Arm 7", "Income >= $30K")

rownames(sum2) <- NULL

sum2 <- sum2 %>%  dplyr::select(names, everything())

round_3 <- function(x){round(x,3)}
sum2[,2:5] <- lapply(sum2[,2:5], round_3)

colnames(sum2) <-c("Model Term", "Estimate", "Std. Error", "T-value", "P-value") 

kbl(sum2, booktabs = T, caption = "") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))

```

\newpage 

# Appendix: 12.2 


```{r}
# look at the correlation between age and age^2 
ggplot(data = infants_f, 
       aes(x = m_age, 
           y = m_age^2 )) + geom_point() + 
  
  stat_smooth(method = "lm", se = F, aes(color = "Fitted Regression Line")) + 
  scale_color_manual(values = c("Fitted Regression Line" = "red")) + 
  
  xlab("Age") + 
  ylab("Age Squared") + 
  ggtitle(paste("Correlation: ", round(cor(infants_f$m_age, infants_f$m_age^2),4))) + 
  theme_minimal()


# now apply centering: 

infants_f$m_age_centered <- with(infants_f, m_age - mean(m_age))
infants_f$gest_weeks_centered <- with(infants_f, gest_weeks - mean(gest_weeks))

ggplot(data = infants_f, 
       aes(x = m_age_centered, 
           y = m_age_centered^2 )) + geom_point() +  
  
  stat_smooth(method = "lm", se = F, aes(color = "Fitted Regression Line")) + 
  scale_color_manual(values = c("Fitted Regression Line" = "red")) + 
  
  xlab("Age") + 
  ylab("Age Squared") + 
  ggtitle(paste("Correlation: ", round(cor(infants_f$m_age_centered, infants_f$m_age_centered^2),4))) + 
  theme_minimal()

```


```{r}
ggplot(data = infants_f, 
       aes(x = m_age, 
           y = birth_w)) + geom_point() + 
  
  stat_smooth(se = F, aes(color = "Smooth Trend Line")) + 
  stat_smooth(se = F, method = "lm", aes(color = "Regression Line")) + 
  
  scale_color_manual(values = c("Smooth Trend Line" = "blue", 
                                "Regression Line" = "red")) + 
  xlab("Mother's Age") + 
  ylab("Infant's Birth Weight") + 
  ggtitle(paste("Correlation Between Mother's Age  \n and Infant's Birth Weight: ", 
                round(cor(infants_f$m_age, 
                          infants_f$birth_w), 4))) + 
  theme_minimal()
  
ggplot(data = infants_f, 
       aes(x = m_age_centered, 
           y = birth_w)) + geom_point() + 
  
  stat_smooth(se = F, aes(color = "Smooth Trend Line")) + 
  stat_smooth(se = F, method = "lm", aes(color = "Regression Line")) + 
  
  scale_color_manual(values = c("Smooth Trend Line" = "blue", 
                                "Regression Line" = "red")) + 
  xlab("Mother's Age") + 
  ylab("Infant's Birth Weight") + 
  ggtitle(paste("Correlation Between Centered Mother's Age  \n and Infant's Birth Weight: ", 
                round(cor(infants_f$m_age_centered, 
                          infants_f$birth_w), 4))) + 
  theme_minimal()

```












