---
title: "Homework 6"
author: "Denis Ostroushko"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    extra_dependencies: ["float"]
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo = F}
knitr::opts_chunk$set(echo = T, message = F, warning = F, fig.pos = "!H")
options(scipen=999)
```

```{r, include=F}
library(tidyverse)
library(kableExtra)
library(readxl)
library(gridExtra)
library(ggeffects)
library(mltools) # one hot encoding outside of caret package 
library(data.table) # need this for mltools to work 
```

```{r, eval=F}
library(tidyverse)
library(kableExtra)
library(readxl)
library(gridExtra)
library(ggeffects)
library(mltools) # one hot encoding outside of caret package 
library(data.table) # need this for mltools to work 
```


# 12.2 

```{r}
infants <- readxl::read_xls('/Users/denisostroushko/Desktop/UofM MS/MS Fall 2022/Puhb 7405/Data Sets/Infants.xls')

colnames(infants) <- c("head_c", "length", "gest_weeks", "birth_w", "m_age", "toxemia")

# process the data and keep variables for analysis 

infants_f <- infants %>% 
  select(birth_w, gest_weeks, m_age)

```

```{r, echo = F}
# look at the correlation between age and age^2 
ggplot(data = infants_f, 
       aes(x = m_age, 
           y = m_age^2 )) + geom_point() + 
  
  stat_smooth(method = "lm", se = F, aes(color = "Fitted Regression Line")) + 
  scale_color_manual(values = c("Fitted Regression Line" = "red")) + 
  
  xlab("Age") + 
  ylab("Age Squared") + 
  ggtitle(paste("Correlation: ", round(cor(infants_f$m_age, infants_f$m_age^2),4))) + 
  theme_minimal()

```

## 12.2 - A

### Model Specifications and T-tests

```{r, echo = F}
ggplot(data = infants_f, 
       aes(x = m_age, 
           y = birth_w)) + geom_point() + 
  
  stat_smooth(se = F, aes(color = "Smooth Trend Line")) + 
  stat_smooth(se = F, method = "lm", aes(color = "Regression Line")) + 
  
  scale_color_manual(values = c("Smooth Trend Line" = "blue", 
                                "Regression Line" = "red")) + 
  xlab("Mother's Age") + 
  ylab("Infant's Birth Weight") + 
  ggtitle(paste("Correlation Between Mother's Age  \n and Infant's Birth Weight: ", 
                round(cor(infants_f$m_age, 
                          infants_f$birth_w), 4))) + 
  theme_minimal()

```

Model specification: 

$$\Large E[Y] = \hat \beta_0 + \hat \beta_1 * Gestional \ Weeks + \hat \beta_2 * Mother's \ Age + \hat \beta_3 * Mother's \ Age ^ 2$$
Model Summary 

```{r, echo = F}
inf_lm <- lm(birth_w ~ gest_weeks + m_age + I(m_age^2), data = infants_f)

sum_data <- data.frame(summary(inf_lm)$coefficients)

sum_data$names <- c("Intercept", "Gestational Weeks", "Mother's Age", "Mother's Age Squared")

rownames(sum_data) <- NULL

sum_data <- sum_data %>%  dplyr::select(names, everything())

round_3 <- function(x){round(x,3)}
sum_data[,2:5] <- lapply(sum_data[,2:5], round_3)

colnames(sum_data) <-c("Model Term", "Estimate", "Std. Error", "T-value", "P-value") 

sum_data %>% 
  kbl(booktabs = T, align = c('l', 'c','c','c','c')) %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))

```

Comemnts on Model summary: 

* R and Adjusted R: `r round(summary(inf_lm)$r.squared,4)` 
`r round(summary(inf_lm)$adj.r.squared,4) `

* Coefficients for Age and Age ^ 2

### Evaluate Extra Sum of Squares

Focus: Evaluate SSR(Age^2 | Gest, Age)

```{r, echo = F}
anv <- data.frame(anova(inf_lm) )

anv$names <- c("Gestational Weeks", "Mother's Age", "Mother's Age Squared", "Residuals")
rownames(anv) <- NULL
anv <- anv %>% select(names, everything())

colnames(anv) <- c("Model Term", "DF", "SS", "MS", "F-statistic", "P(F* > F)")

anv[,length(anv)] <- round(anv[,length(anv)], 4)

anv %>% 
  kbl(booktabs = T, align = c('l', rep('c', length(anv)-1)) ) %>% 
  kable_styling(latex_options = c("striped", "HOLD_position")) %>% 
  row_spec(nrow(anv)-1, hline_after = T)

```

```{r, echo = F}
# R^2 (m age | gest weeks )
lm_2 <- lm(birth_w ~ gest_weeks + m_age, data = infants_f)
ssr_2 <- sum( (lm_2$fitted.values - mean(infants_f$birth_w) )^2 )

# R^2 (m age ^2  | m age, gest weeks )
lm_3 <- lm(birth_w ~ gest_weeks + m_age + I(m_age^2), data = infants_f)

sse_2 <- sum(lm_2$residuals^2)
ssr_3 <- sum( (lm_3$fitted.values - mean(infants_f$birth_w) )^2 )

Rs_3 <- (ssr_3 - ssr_2) /sse_2

```

* Extra SS

* Extra R^2 

* Connection with the t-test 

### Visualize Model Effects

```{r, echo = F }
mydf <- ggpredict(inf_lm, terms = "m_age")

mean_name <- paste0("Birth Weight Mean Value: ", round(mean(infants_f$birth_w)))

ggplot(mydf, aes(x, predicted)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = .1) + 
  theme_minimal() + 
  geom_hline( 
             aes(yintercept = mean(infants_f$birth_w),color = mean_name)
             ) + 
  guides(color=guide_legend(title="Additional Elements: ")) + 
  theme(
    legend.position = "bottom"
  ) + 
  
  ylab("Predicted Values of Birth Weight") + 
  xlab("Mother's Age") + 
  ggtitle("Model Estiamted Effects of Mother's Age on \n Infant's Birth Weight")
  

```

* Comment on Standard Error and fit, we can fit a line with slope = +

### Interpretation of Mother's Age Coefficients 

From google, interpretation of the quadratic coefficient: 

"
A positive quadratic coefficient causes the ends of the parabola to point upward. A negative quadratic coefficient causes the ends of the parabola to point downward. The greater the quadratic coefficient, the narrower the parabola. The lesser the quadratic coefficient, the wider the parabola."

https://stats.stackexchange.com/questions/108657/how-to-interpret-coefficients-of-x-and-x2-in-same-regression

It may be useful to describe the effect of a unit change at some low value, some high value and somewhere in between.

## 12.2 - B

Correlation Transformation for variables $Y, X_1, ..., X_{p-1}$, denoted by $V$: 

$$\Large V^* = \frac{1}{\sqrt{n - 1}} \times \Big (\frac{V - \bar V}{sd(V)} \Big )$$
```{r}

correlation_transformation <- 
  function(X, n = nrow(infants_f_cor_tr)){
    
    1/(sqrt(n - 1)) * (X - mean(X))/sd(X)
    
  }

infants_f$m_age_sq <- infants_f$m_age^2
infants_f_cor_tr <- infants_f

infants_f_cor_tr <- data.frame(lapply(infants_f_cor_tr, correlation_transformation))
```

```{r, echo = F}
cor_tr_lm <- lm(birth_w ~ . , data = infants_f_cor_tr)
sum2 <- data.frame(summary(cor_tr_lm)$coefficients)

sum2$names <- c("Intercept", "Gestational Weeks", "Mother's Age", "Mother's Age Squared")

rownames(sum2) <- NULL

sum2 <- sum2 %>%  dplyr::select(names, everything())

round_3 <- function(x){round(x,3)}
sum2[,2:5] <- lapply(sum2[,2:5], round_3)

colnames(sum2) <-c("Model Term", "Estimate", "Std. Error", "T-value", "P-value") 

kbl(sum_data, booktabs = T, caption = "Original Scale Regression Estimates") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))

kbl(sum2, booktabs = T, caption = "Correlation  Transformation Regression Estimates") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))

```

* intercept is zero as expected in corr transformed 

* P-values are different for m age 

* Same conclusions apply 

## 12.2 - C

Transformation back to the original scale: 

For variables $X_1, ..., X_{p-1}$: 

$$\Large \hat \beta_i =  \hat \beta^*_i \times \frac{sd(Y)}{sd(X_i)}$$
```{r}
transform_back <-
  function(Beta_star, s_x, s_y){
    Beta_star * (s_y / s_x)
  }

S_Y <- sd(infants_f$birth_w)
```

Hid code to prepare the table. 

```{r, echo = F}
check_res <- 
  data.frame(
    names = c("Gestation Weeks", "Mother's Age", "Mother's Age Squared"), 
    betas = c(transform_back(sum2[2,2], s_x = sd(infants_f$gest_weeks), s_y = S_Y),
              transform_back(sum2[3,2], s_x = sd(infants_f$m_age), s_y = S_Y),
              transform_back(sum2[4,2], s_x = sd(infants_f$m_age_sq), s_y = S_Y)
              ), 
    ci_l = c(transform_back(confint(cor_tr_lm)[2,1], s_x = sd(infants_f$gest_weeks), s_y = S_Y),
              transform_back(confint(cor_tr_lm)[3,1], s_x = sd(infants_f$m_age), s_y = S_Y),
              transform_back(confint(cor_tr_lm)[4,1], s_x = sd(infants_f$m_age_sq), s_y = S_Y)
              ),
    ci_h = c(transform_back(confint(cor_tr_lm)[2,2], s_x = sd(infants_f$gest_weeks), s_y = S_Y),
              transform_back(confint(cor_tr_lm)[3,2], s_x = sd(infants_f$m_age), s_y = S_Y),
              transform_back(confint(cor_tr_lm)[4,2], s_x = sd(infants_f$m_age_sq), s_y = S_Y)
              )
  )

```


recall the the original model with the transformed variables was called `inf_lm`. Used it for Extra SS,
t-tests and model effects. We can obtain standard errors and confidence intervals for the estimates
to compare with the transformation back from the correlation transformation procedure. 

```{r}
conf <- data.frame(confint(inf_lm)) # just the confidence intervals 
conf <- cbind(coefficients(inf_lm), conf )
```

```{r, echo=F}
conf$names <- c("Intercept", "Gestation Weeks", "Mother's Age", "Mother's Age Squared")
rownames(conf) <- NULL

conf <- conf %>% select(names, everything())

colnames(conf) <- c("Model Term", "Coefficient", "95% C.I. Lower Bound", "95% C.I. Upper Bound")

round_3 <- function(x){round(x,3)}
conf[,2:4] <- sapply(conf[, 2:4], round_3)

conf %>% filter(`Model Term` != "Intercept" ) %>% 
  kbl(align = c('l', rep('c', length(conf)-1)), booktabs =  T, 
      caption = "Original Model Estiamtes and C.I.") %>% 
  kable_styling(latex_options = c("HOLD_posiiton", "striped")) 

```

```{r, echo = F}

colnames(check_res) <- c("Model Term", "Coefficient", "95% C.I. Lower Bound", "95% C.I. Upper Bound")

check_res[,2:4] <- lapply(check_res[,2:4], round_3)

check_res %>% filter(`Model Term` != "Intercept" ) %>% 
  kbl(align = c('l', rep('c', length(conf)-1)), booktabs =  T, 
      caption = "Estimaes obtained via \n Back-Trnasformation and C.I.") %>% 
  kable_styling(latex_options = c("HOLD_posiiton", "striped")) 
```

so we can use linear transformations 
good to know 

# 13.4 

```{r}
cig <- read_xlsx('/Users/denisostroushko/Desktop/UofM MS/MS Fall 2022/Puhb 7405/Data Sets/E-CID-3.xlsx')

cig$Y1 <- with(cig, log(NNAL_vt4_creat / NNAL_vt0_creat))
cig$Y2 <- with(cig, log(TNE_vt4_creat / TNE_vt0_creat))

cig <- cig %>% 
  select(Y1, Y2, arm, age, gender, white, educ2, income30, FTND)

colnames(cig)[length(cig)] <- "ftnd"
```

## 13.4 - A

* Arm will result in 4 -1 variables 

* Age is untouched 

* FTND is treated as continuous 

* Others need to be converted to factor variables 

```{r}
cig <- cig %>% select(
  Y1, Y2, age, arm, gender, white, educ2, income30, ftnd
)

cig$arm <- as.factor(cig$arm)

cig <- data.frame(one_hot(as.data.table(cig))) %>% select(-arm_5)

cig[,6:length(cig)] <- lapply(cig[,6:length(cig)], as.factor)
```

```{r}
n_unique <- function(x){length(unique(x))}

meta_data <- 
  
  data.frame(
    class = sapply(cig, class), 
    n_unique = sapply(cig, n_unique)
  )
```

```{r, echo = F}

meta_data$name <- rownames(meta_data)

rownames(meta_data) <- NULL

meta_data <- meta_data %>% filter(!(name %in% c("Y1", "Y2")))

meta_data <- 
  meta_data %>% select(name, everything())
  
colnames(meta_data) <- c("Predictors", "Assigned Class", "N of Unique Values")

meta_data %>% 
  kbl(caption = "Sumamry of Covariates",
       booktabs = T, align = c("l", 'c', 'c')) %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))

nrow(meta_data)
```

## 13.4 - B

### Regression on Y1 

```{r}
y1_lm1 <- lm(Y1 ~ ., data = cig %>% select(-Y2))
summary(y1_lm1)
```

```{r}
sum2 <- data.frame(summary(y1_lm1)$coefficients)

sum2$names <- c("Intercept", "Arm 6", "Arm 7", "Arm 8", 
                "Age", "Gender", "White")

rownames(sum2) <- NULL

sum2 <- sum2 %>%  dplyr::select(names, everything())

round_3 <- function(x){round(x,3)}
sum2[,2:5] <- lapply(sum2[,2:5], round_3)

colnames(sum2) <-c("Model Term", "Estimate", "Std. Error", "T-value", "P-value") 

kbl(sum_data, booktabs = T, caption = "Original Scale Regression Estimates") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))
```

* Bonferroni Adjustments 

* HOLM Adjustments 

* Hochberg Adjustments

### Regression on Y2 

```{r}
y2_lm1 <- lm(Y2 ~ ., data = cig %>% select(-Y1))
summary(y2_lm1)
```

* Bonferroni Adjustments 

* HOLM Adjustments 

* Hochberg Adjustments

## 13.4 - C

### Step Wise Regression on Y1 

### Step Wise Regression on Y1 

# Appendix: 12.2 


```{r}
# look at the correlation between age and age^2 
ggplot(data = infants_f, 
       aes(x = m_age, 
           y = m_age^2 )) + geom_point() + 
  
  stat_smooth(method = "lm", se = F, aes(color = "Fitted Regression Line")) + 
  scale_color_manual(values = c("Fitted Regression Line" = "red")) + 
  
  xlab("Age") + 
  ylab("Age Squared") + 
  ggtitle(paste("Correlation: ", round(cor(infants_f$m_age, infants_f$m_age^2),4))) + 
  theme_minimal()


# now apply centering: 

infants_f$m_age_centered <- with(infants_f, m_age - mean(m_age))
infants_f$gest_weeks_centered <- with(infants_f, gest_weeks - mean(gest_weeks))

ggplot(data = infants_f, 
       aes(x = m_age_centered, 
           y = m_age_centered^2 )) + geom_point() +  
  
  stat_smooth(method = "lm", se = F, aes(color = "Fitted Regression Line")) + 
  scale_color_manual(values = c("Fitted Regression Line" = "red")) + 
  
  xlab("Age") + 
  ylab("Age Squared") + 
  ggtitle(paste("Correlation: ", round(cor(infants_f$m_age_centered, infants_f$m_age_centered^2),4))) + 
  theme_minimal()

```


```{r}
ggplot(data = infants_f, 
       aes(x = m_age, 
           y = birth_w)) + geom_point() + 
  
  stat_smooth(se = F, aes(color = "Smooth Trend Line")) + 
  stat_smooth(se = F, method = "lm", aes(color = "Regression Line")) + 
  
  scale_color_manual(values = c("Smooth Trend Line" = "blue", 
                                "Regression Line" = "red")) + 
  xlab("Mother's Age") + 
  ylab("Infant's Birth Weight") + 
  ggtitle(paste("Correlation Between Mother's Age  \n and Infant's Birth Weight: ", 
                round(cor(infants_f$m_age, 
                          infants_f$birth_w), 4))) + 
  theme_minimal()
  
ggplot(data = infants_f, 
       aes(x = m_age_centered, 
           y = birth_w)) + geom_point() + 
  
  stat_smooth(se = F, aes(color = "Smooth Trend Line")) + 
  stat_smooth(se = F, method = "lm", aes(color = "Regression Line")) + 
  
  scale_color_manual(values = c("Smooth Trend Line" = "blue", 
                                "Regression Line" = "red")) + 
  xlab("Mother's Age") + 
  ylab("Infant's Birth Weight") + 
  ggtitle(paste("Correlation Between Centered Mother's Age  \n and Infant's Birth Weight: ", 
                round(cor(infants_f$m_age_centered, 
                          infants_f$birth_w), 4))) + 
  theme_minimal()

```












