---
title: "Homework 2"
author: "Denis Ostroushko"
date: '`r as.Date(Sys.Date())`'
output: html_document
editor_options:
  markdown:
    wrap: 72
---

<!--
Set up packages and fucntions 
--> 

```{R, warning = F, message = F}

# load packages 

library(tidyverse)
library(readxl)

```

```{r read initial data 4_2, message=F, warning=F}

vc_data <- 
  read_xlsx("/Users/denisostroushko/Desktop/UofM MS/MS Fall 2022/Puhb 7405/Data Sets/VitalCapacity.xlsx")
vc_data <- na.omit(vc_data)

colnames(vc_data) <- c("age",  "xvc100")

```

# 4.2

<!-- 

Latex tips: use single dollar sign in line to make equations and greek letters 
use double dollar sign to make equations centered and make them stand out in the rendered output 

--> 

## 4.2 - A

First need $b_1$, estimate for $\beta_1$
Formula:

$$ \Large b_1 = \frac{\Sigma (X_i - \bar X)(Y_i - \bar Y)}{\Sigma (X_i - \bar X)^2} $$

Code is below 

```{r b1_4_2_estimate}

b1 <- 
  sum(
    (vc_data$age - mean(vc_data$age)) * #(x_i - x_bar)
        (vc_data$xvc100 -mean(vc_data$xvc100)) #(y_i - y_bar)
    ) / 
  sum((vc_data$age - mean(vc_data$age))^2) # squared error of X

```

Estimate for $\beta_1$ = $\hat \beta_1$ = $b_1$ = `r b1`

Now we can obtain $b_0$

Formula: 

$$\Large \frac{1}{n} (\Sigma Y_i - b_1 \times \Sigma X_i) = \bar Y - b_1 \times \bar X$$

Code to estimate is below: 

```{r b_0_4_2_estimate}

b0 <- 
  mean(vc_data$xvc100) - b1 * mean(vc_data$age)

```

Estimate for $\beta_0$ = $\hat \beta_0$ = $b_0$ = `r b0`

We can check our answer by creating a linear model: 

```{R 4_2_a_linear model }

lm_4_2_a <- lm(xvc100 ~ age, data = vc_data)

sum <- summary(lm_4_2_a)

sum

b0_model <- sum$coefficients[1]
b1_model <- sum$coefficients[2]

```

From the summary of the model we can see that $b_0$ = `r b0_model` and $b_1$ = `r b1_model`

Difference in two estimates for $b_1$ is `r round(b1 - b1_model,4)`, rounded to 4 decimal points.
Difference in two estimates for $b_0$ is `r round(b0 - b0_model,4)`, rounded to 4 decimal points.

## 4.2 - B

```{r plot, message = F, warning = F}

#save estimated data - Y_hat from the mdoel - to the data set for convinient plotting 

vc_data$y_hat <- predict(lm_4_2_a, vc_data)

min_round_bound <- round(log10(min(vc_data$xvc100))) * 10^round(log10(min(vc_data$xvc100)))
max_round_bound <- round(log10(max(vc_data$xvc100))) * 10^round(log10(max(vc_data$xvc100)))

ggplot(
  data = vc_data, 
  aes(x = age, y = xvc100)
) + 
  geom_point() + 
  geom_line(data = vc_data, aes(x = age, y = y_hat, colour = "Estimated Regression Line"), size = 1) + 
  geom_smooth(aes(colour = "Smooth Trend Line"), se = F) + 
  
  scale_color_manual(name = "Line Types", values = c("Estimated Regression Line" = "red", "Smooth Trend Line" = "blue")) + 
  
  scale_y_continuous(breaks = seq(from = min_round_bound, to = max_round_bound, by = 25)) + 
  ylab("Lung Vital Capacity") + 
  xlab("Age") + 
  ggtitle("Relationship Between Age in Years and Lung Capacity") + 
  theme_minimal()

```

## 4.2 - C

to obtain this estimate we simply need to plug in the value of $X = 35$ into our regression equation

Estimated value is `r b0 + b1*35`

We can also obtain a confidence interval for the mean response level when $X_h = 35$

First, we want a standard error, so we need MSE (Mean Squared Error, obtain from residuals), deviation of $X_h$ from the mean, $\bar X$, and total variance of $X$

We will also obtain a coefficient from the t distribution, at 95% confidence level and `r nrow(vc_data) - 2` degrees of freedom

```{r}

## se^2 formula: mse (1/n + (x-x_var)^2 / ssx)

X_h <- 35 # desired X level for prediction 

df <- nrow(vc_data - 2) # degrees of freedom for studentized T distribution 

mse <- sum((vc_data$xvc100 - vc_data$y_hat)^2)/df # mean squared error term from Y -> residals -> ovserved minus predicted 
ssx <- sum((vc_data$age - mean(vc_data$age))^2) # sum of squared X deviations 

se_sq <- mse * (1/nrow(vc_data) + (X_h - mean(vc_data$age))^2 / ssx ) #standard error squared 

t_coef <- 
  abs(
    qt(
      1-.05/2, 
      nrow(vc_data) - 2
      )
  ) # coefficient T 

est <- b0 + b1*X_h # point estimate of E[Y_h]

ci_low <- est - sqrt(se_sq) * t_coef # lower bound of confidence interval 
ci_high <- est + sqrt(se_sq) * t_coef # upper bound of confidence interval 
```

```{r}
# predict mean response using R function 

predict(lm_4_2_a, data.frame(age = X_h), interval = "confidence")
```



## 4.2 - D

Mean response -> average increase in response - VC function, when X - age - is increased by 1 unit, i.e. when person gets one year odler 
When age increases by 1 year then VC decreases by `r abs(b1)` . 
Since the relationship is linear, when a person gets 10 years older, VC descreases by `r 10 * abs(b1)`

# 5.3

```{r}
cig <- read_xls("/Users/denisostroushko/Desktop/UofM MS/MS Fall 2022/Puhb 7405/Data Sets/Cigarettes.xls")

original_names <- colnames(cig)

colnames(cig) <- c("age", "gender", "cpd", "carbon_mono", "cotinine", "nnal")

```

## 5.3 - A

We need to estimate average response when $X_h = 30$, 30 cigarettes per day. 

We showed how to estimate model parameters, average reposnse level and confidence interval by hand in 4.2, so we will use R functions to 
get estimates for interpretations 

```{r}
lm_5_3 <- lm(cotinine ~ cpd, data = cig)
summary(lm_5_3)

predict(lm_5_3, data.frame(cpd = 30), interval = "confidence")

b1 <- summary(lm_5_3)$coefficients[2]
b0 <- summary(lm_5_3)$coefficients[1]
```

## 5.3 - B 

Here we need a new method of getting a confidence interval. Begin by derviing a standard error for a signle prediction 

$$se(Y_{h (new)})^2 = MSE * [1 + \frac{1}{n} + \frac{(X_{h (new)} - \bar X)^2}{\Sigma (X_i - \bar X)^2}]$$

```{r}

X_h <- 30

n <- nrow(cig)

mse <- sum(lm_5_3$residuals^2)/(n-2)

ssx <- sum( (cig$cpd - mean(cig$cpd))^2 )

se_y_h_new <- 
  sqrt(
    mse * (1 + 1/n + (X_h - mean(cig$cpd))^2 / ssx)
  )

t_coef <- qt(1 - .05/2, n - 2)

ci_lower <- (b1 * X_h + b0) - se_y_h_new * t_coef
ci_upper <- (b1 * X_h + b0) + se_y_h_new * t_coef
```

```{r}
ci_lower 
ci_upper 

predict(lm_5_3, data.frame(cpd = 30), interval = "prediction")

```

## 5.3 - C

```{r}
cig$resid <- lm_5_3$residuals

ggplot(data = cig, 
       aes(x = cpd, y = resid)) + 
  geom_point() + 
  geom_smooth(aes(color = "Smooth Trend Line"), se = F, size = 1) + 
  stat_smooth(aes(color = "Regression Line") , se = F,  size = 1, method = "lm") + 
  
  xlab("Cigarettes per Day") + 
  ylab("Residuals ") + 
  ggtitle( "Plot of X values against Residuals "
  ) + 
  
  scale_color_manual(name = "Line Types", values = c("Smooth Trend Line" = "blue", "Regression Line" = "red"))
```

```{r}
summary(cig$resid)
```

average residual value: `r mean(cig$resid)`

## 5.3 - D

```{r}
anova(lm_5_3)

f_cutoff <- qf(1-0.05, 1, nrow(cig) - 2)
f_cutoff
```
