---
title: "Homework 10"
author: "Denis Ostroushko"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    extra_dependencies: ["float"]
editor_options: 
  markdown: 
    wrap: 72
---



```{r, echo = F}
knitr::opts_chunk$set(echo = F, message = F, warning = F, fig.pos = "!H", fig.height=4, fig.width=7, fig.align='center')
options(scipen=999)
```

```{r load all packages from the master file , include=F}
source('/Users/denisostroushko/Desktop/UofM MS/MS Fall 2022/Puhb 7405/Package master list .R')
```

# 20.2

```{r}

censored_case = 
  c(0,0,1,1,0,0,0 ,0 ,0,0,0,0,0,0,0,0,0,0,0,1,0,0
)

time_case = 
  c(30,67,79,82,95,148,170 ,171 ,176,193,200,221,243,261,262,263,399,414,446,446,464,777 
)

time_control = 
  c(57,58,74,79,89,98,101,104,110,118,125, 132, 154, 159,188, 203, 257, 257, 431, 461, 497, 723, 747,1313,2636
)

censored_control <- rep(0, length(time_control))

final_data <- 
  rbind(
    data.frame(
      time = time_case, 
      censored = censored_case, 
      type = "case"
    ), 
    data.frame(
      time = time_control, 
      censored = censored_control, 
      type = "control"
    )
  )

final_data$status <- 1 - final_data$censored 
```

### 20.2 - 1

In this section we derive KM survival estimates for each treatment arm and present them in the table. Each table will have 
all detail that goes into the KM estimates: subjects at risk, number of events at time point $i$, instantaneous hazard, and 
survival probability estimate. 

This list describes each column of the final table that, and provides formulas to be used in calculations: 

1. Number of subjects at risk is the number of remaining subjects at time $i$ who were not censored or did not have an event, 
denote this quantity as $n_i$

2. Number of events will occurring at time $i$ will be used for calculation of hazard, denote as $d_i$

3. Hazard at time $i$ is $\large \lambda_i = \frac{d_i}{n_i}$. We need $\large 1 - \lambda_i$ as the instantaneous probability 
  to survive

4. Survival Probability at time $i$ depends on the time period: 
  + if it is the first available time period, then $\large S(1) = 1 - \lambda_1$
  + Otherwise, $\large S(i) = (1 -\lambda_i) \times S(i-1)$

```{r}

cases <- final_data %>% filter(type == "case")

cases %>% 
  group_by(time) %>% 
  summarize(n = n(), 
            events = sum(1-censored)) %>% 
  arrange(-time) %>% 
  mutate(n_i = cumsum(n)) %>% 
  arrange(time) %>% 
  mutate(inst = 1-events/n_i) -> cases_f

cases_f$S <- cases_f$inst

for(i in 2:nrow(cases_f)){
  cases_f$S[i] <- cases_f$inst[i] * cases_f$S[i-1]
}
 
cases_f <- cases_f %>% select(time, n_i, events, inst, S)
```

Clearly, in the case of the first time interval recorded in our data at time = `r min(cases$time)` we set survival probability equal to the instantaneous probability of surviving. Technically, we can express it as
a conditional probability, where survival probability before the start of observation is set to 1. Noone simply experienced the 
event yet, so everybody "survived" to that point. 

At time = `r min(cases[cases$time != min(cases$time),]$time)`, survival probability is 
now truly a conditional probability, expressed as $\large S_1 \times (1-\lambda_2)$, which is 
`r round(cases_f[1, ]$S, 4)` $\times$ 
`r round(cases_f[2, ]$inst, 4)`

Full Table is given below: 

```{r}
cases_f %>% 
  kbl(booktabs = T, align = 'c', 
      col.names = c("Time", "At Risk: n_i", "Events: d_i", "Instantaneous survival Probability: 1 - d_i/n_i", 
                    "KM Survival Estimate"), 
      caption = "Kaplan Meier Survival Estiamtes for 'RoRx+5-Fu' Case Group ") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position")) %>% 
  column_spec(4, width = "4cm")
```

We also replicate this method to provide a table for controls: 

```{r}
controls <- final_data %>% filter(type == "control")

controls %>% 
  group_by(time) %>% 
  summarize(n = n(), 
            events = sum(1-censored)) %>% 
  arrange(-time) %>% 
  mutate(n_i = cumsum(n)) %>% 
  arrange(time) %>% 
  mutate(inst = 1-events/n_i) -> controls_f

controls_f$S <- controls_f$inst

for(i in 2:nrow(controls_f)){
  controls_f$S[i] <- controls_f$inst[i] * controls_f$S[i-1]
}
 
controls_f <- controls_f %>% select(time, n_i, events, inst, S)

controls_f %>% 
  kbl(booktabs = T, align = 'c', 
      col.names = c("Time", "At Risk: n_i", "Events: d_i", "Instantaneous survival Probability: 1 - d_i/n_i", 
                    "KM Survival Estimate"), 
      caption = "Kaplan Meier Survival Estiamtes for Control Group ") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position")) %>% 
  column_spec(4, width = "4cm")
```

### 20.2 - 2

In this section we put the data from two tables on the same graphs. We also add median survival time for both groups. 
From slide #5 in Lecture file #21 it appears that the median time is defined as the time point when the value of survival function 
$S(t)$ drop below 0.5 for the first time.  We can't just take the median of observed survival times due to cencoring and 
incomplete data that it imposes. 

```{r}

case_median_t <- min(cases_f[cases_f$S < .5, ]$time)
cont_median_t <- min(controls_f[controls_f$S < .5, ]$time)

```

Thus, as we can see in the tables, median survival time for cases who are subject to treatment is `r case_median_t` days, while 
it is `r cont_median_t` days for controls.

```{r}
fit <- survfit(Surv(time, status) ~ type, data = final_data)
plot <- 
  ggsurvplot(fit, data = final_data, surv.median.line = "hv") + 
  ggtitle(paste("Kaplan Meier Survival Curves. \n   Cases Median Survival Time: ", case_median_t, 
                "days. \n   Controls Median Survival Time: ", cont_median_t, " days.")) 

plot$plot

# plot$plot + 
#   geom_vline(aes(xintercept = case_median_t, color = "Case median survival time"), 
#                   size = 1, linetype = "dashed") + 
#   geom_vline(aes(xintercept = cont_median_t, color = "Control median survival time"), 
#                   size = 1, linetype = "dashed")
```

In appears that we have the phenomenon of crossing survival curves. That means that he hazards are not proportional, and 
if we are to analyze these curves and compare survival times we would need parametric methods that can handle different scale and 
shape parameters for the two groups. 

Judging from the location of the median survival times, cases' survival time should be higher, "on average" by around 
`r case_median_t - cont_median_t` days. 

That means that the drug we are researching may be effective at extending survival time of patients. 
However, as we noted, at some point survival curves cross. Which means that the drug must have some adverse effects on the 
subjects after some prolonged period of time. 

Alternatively, the drug potency may simply decrease over time and may no longer deter the unfavorable outcomes. 

### 20.2 - 3

Output of the Cox Proportional Hazard Model is given below: 

```{r}
res.cox <- coxph(Surv(time, status) ~ type, data = final_data)

summary(res.cox)$coefficient

```

Exponentiated coefficient is `r round(summary(res.cox)$coefficient[2], 4)` meaning that the controls' instantaneous  hazard 
at any time is `r paste((round(summary(res.cox)$coefficient[2], 4) -1)*100, "%") ` higher at any time, when compared with cases. 

However, low Wald Z statistic, high standard error, and low p-value indicate that this difference is not statistically significant.

We can also obtain fitted survival curves from the Cox Proportional Hazards Regression Model. 
As we can see, the two curves are hard to distinguish. 

```{r}
ggadjustedcurves(res.cox, variable = 'type') + 
  ggtitle("Fitted Survival Curves")
```

# 21.1

```{r}

placebo_time <- c(
  1,2,2,2,3,4,4,5,5,8,8,8,8,11,11,12,12,15,17, 22,23)

placebo_censored <- rep(0, length(placebo_time))

placebo_status <- 1- placebo_censored

case_time <- c(
  6,6,6,7,10,13,16,22,23,6,9,10,11,17,19,20,25,32,32,34,35)

case_censored <- c(
  0,0,0,0,0,0,  0, 0, 0, 1, 1, 1,  1,  1,  1,  1,  1,  1,  1,  1,  1)

case_status <- 1 - case_censored

final_data <- 
  rbind(
    data.frame(
      time = placebo_time, 
      censored = placebo_censored, 
      status = placebo_status, 
      type = "placebo"
    ),
    data.frame(
      time = case_time, 
      censored = case_censored, 
      status = case_status, 
      type = "cases"
    )
  ) %>% 
  arrange(type, time)

```

### 21.1 - 1

Using the same methods we used in problem #20, we provide the two tables, one for each group, with KM survival estimates  at 
each point in time. 

```{r}
controls <- final_data %>% filter(type == "placebo")

controls %>% 
  group_by(time) %>% 
  summarize(n = n(), 
            events = sum(1-censored)) %>% 
  arrange(-time) %>% 
  mutate(n_i = cumsum(n)) %>% 
  arrange(time) %>% 
  mutate(inst = 1-events/n_i) -> controls_f

controls_f$S <- controls_f$inst

for(i in 2:nrow(controls_f)){
  controls_f$S[i] <- controls_f$inst[i] * controls_f$S[i-1]
}
 
controls_f <- controls_f %>% select(time, n_i, events, inst, S)

controls_f %>% 
  kbl(booktabs = T, align = 'c', 
      col.names = c("Time", "At Risk: n_i", "Events: d_i", "Instantaneous survival Probability: 1 - d_i/n_i", 
                    "KM Survival Estimate"), 
      caption = "Kaplan Meier Survival Estiamtes for Placebo Group ") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position")) %>% 
  column_spec(4, width = "4cm")
```

```{r}
cases <- final_data %>% filter(type == "cases")

cases %>% 
  group_by(time) %>% 
  summarize(n = n(), 
            events = sum(1-censored)) %>% 
  arrange(-time) %>% 
  mutate(n_i = cumsum(n)) %>% 
  arrange(time) %>% 
  mutate(inst = 1-events/n_i) -> cases_f

cases_f$S <- cases_f$inst

for(i in 2:nrow(cases_f)){
  cases_f$S[i] <- cases_f$inst[i] * cases_f$S[i-1]
}
 
cases_f <- cases_f %>% select(time, n_i, events, inst, S)

cases_f %>% 
  kbl(booktabs = T, align = 'c', 
      col.names = c("Time", "At Risk: n_i", "Events: d_i", "Instantaneous survival Probability: 1 - d_i/n_i", 
                    "KM Survival Estimate"), 
      caption = "Kaplan Meier Survival Estiamtes for 6-MP Cases Group ") %>% 
  kable_styling(latex_options = c("striped", "HOLD_position")) %>% 
  column_spec(4, width = "4cm")
```

We can also visualize the survival curves for the placebo and cases groups

```{r}

case_median_t <- min(cases_f[cases_f$S < .5, ]$time)
cont_median_t <- min(controls_f[controls_f$S < .5, ]$time)

```

```{r}
fit <- survfit(Surv(time, status) ~ type, data = final_data)

ggsurvplot(fit, data = final_data, , surv.median.line = "hv") + 
  ggtitle(paste("Kaplan Meier Survival Curves. \n   Cases Median Survival Time: ", case_median_t, 
                "Weeks. \n   Placebo Median Survival Time: ", cont_median_t, " Weeks.")) 

```

### 21.1 - 2

Exponential model assumes that the survival probability at time $t$ is given by: $\large S(t) = exp(-\rho t)$

```{r}
fit_exp <- flexsurvreg(Surv(time, status) ~ 1, data = final_data %>% filter(type == "cases"), dist = "exp") 

rho =fit_exp$res[1]
 
rho_l =fit_exp$res[2]
rho_u =fit_exp$res[3]

```

parameter $\large \rho$ is estimated to be `r round(rho, 4)`, with a (`r round(rho_l,4)`, `r round(rho_u,4)`) 95% confidence interval. I suppose that the confidence interval is not equal width around the estimate because we to obtain the estimate 
we use the process of exponentiation. 

```{r}
cases_f_exp <- cases_f
cases_f_exp$S_exp <- exp( - rho * cases_f_exp$time)

```

Table below presents estimated survival rates for cases in the 6-MP trial under the assumption that the survival times are 
distributed exponentially. For comparison, we also keep the Kaplan-Meier estimates in the table. We also present the data 
only for those times $t$ where we observe at least one event, so those time record where we have a record, but only censored 
ones, are not in the table. This is specifically requested in the problem statements in the slides.  

```{r}
cases_f_exp %>% select(time, events, S, S_exp) %>% 
  filter(events > 0) %>% 
  kbl(caption = "KM and Expoenential Model Survival Rates for 6-MP Cases", 
      align = 'c', 
      booktabs = T, 
      col.names = c("Time", "Number of Events", "KM Survival Rate", "Exponential Fit Survival Rate")) %>% 
  kable_styling(latex_options = c("striped", "HOLD_position"))
```

Overall, the two methods produce quite similar estimates, although probabilities from the survival fit appear to be more 
optimistic. Especially as we get to the higher time records, exponential fit probabilities overstate expected survival chance 
quite notably. 

### 21.1 - 3

The plot below visualizes the two survival curves on the same plot. We can see that the two curves align quite well up until time
$t \approx$ 22, where the biggest differences start to occur. The difference occurs because the because KM is fitted to 
our particular sample, while exponential method is not. Overall, either method might not be well suited for prediction of survial 
time or extrapolation of any sort. 

```{r}
plot_data <- 
  cases_f_exp %>% 
  select(time, S, S_exp) %>%
  pivot_longer(
    cols = c("S", "S_exp"), 
    names_to = "names", 
    values_to = "values"
  ) %>% 
  mutate(names = case_when(
    names == "S" ~ "KM Estimate", 
    T ~ "Exponential Fit Estimate"
  ))

ggplot(data = plot_data, 
       aes(x = time, y = values, group = names, color = names)) + 
  geom_point() + geom_step() + 
  theme_minimal() + 
  
  xlab("Time") + 
  ylab("Survival Rates Estiamtes") 

```

The plot below visualizes the relationship between the two sets of survival probabilities. As the problem asks in section (2), 
we remove those time points where no events, and only censoring, occurs. We only have 7 or so data points for this plot, but 
the two sets of probabilities are almost perfectly correlated. Either can be used for estimation of survival time for this sample 
up until time point of 25 weeks or so. 

```{R}
cases_f_exp_e <- cases_f_exp %>% filter(events > 0)

ggplot(data = cases_f_exp_e, 
       aes(x = S, y = S_exp)) + 
  geom_point() + 
  
  geom_smooth(method = "lm", color = "red", se = T) + 
  geom_smooth(color = "blue", se = F) + 
  
  xlab("K-M Survival Estiamte") + 
  ylab("Exponential Fit Survival Estimate") + 
  ggtitle(paste("Correlation between the two set of estimates: ", round(
    cor(cases_f_exp_e$S, cases_f_exp_e$S_exp), 6))
    )+ 
  
  theme_minimal()

```

\newpage

### 21.1 - 4

In order to fit the Weibull time-to-event model we will estimate two parameters. 

```{r}

fit_weib <- flexsurvreg(Surv(time, status) ~ 1, data = final_data %>% filter(type == "cases"), dist = "weibull") 

rho_shape <- fit_weib$coefficients[1]
lambda_scale <- fit_weib$coefficients[2]

```

```{r eval = F, echo = F}
rho_shape
lambda_scale
```

Two parameters: scale - location parameter - $\large \lambda$ = `r round(lambda_scale, 4)`

shape - $\large \rho$ = `r round(rho_shape, 4)`

Survival rate = $S(t) = exp(- (t / \lambda )^\rho)$

We can now use weibull survival model to estimate survival probabilities. As we can see, weibull model is a poor method for 
our sample. 

```{R}

cases_f_weib <- cases_f_exp
cases_f_weib$S_w <- exp( - (cases_f_weib$time / lambda_scale)^rho_shape)

plot_data <- 
  cases_f_weib %>% 
  select(time, S, S_exp, S_w) %>%
  pivot_longer(
    cols = c("S", "S_exp", S_w), 
    names_to = "names", 
    values_to = "values"
  ) %>% 
  mutate(names = case_when(
    names == "S" ~ "KM Estimate", 
    names == "S_w" ~ "Weibull Fit Estimate", 
    T ~ "Exponential Fit Estimate"
  ))

ggplot(data = plot_data, 
       aes(x = time, y = values, group = names, color = names)) + 
  geom_point() + geom_step() + 
  theme_minimal() + 
  
  xlab("Time") + 
  ylab("Survival Rates Estiamtes") 

```

### 21.1 - 5

Again, for correlation plot, we only keep those time point where no events and only censoring occurred. The two sets of probabilities
are highly correlated, however, Weibull tends to severely underestimate the survival probability at any given point in time. 

```{R}

cases_f_weib_e <- cases_f_weib %>% filter(events > 0)

ggplot(data = cases_f_weib_e, 
       aes(x = S, y = S_w)) + 
  geom_point() + 
  
  geom_smooth(method = "lm", color = "red", se = T) + 
  geom_smooth(color = "blue", se = F) + 
  
  xlab("K-M Survival Estiamte") + 
  ylab("Weibull Fit Survival Estimate") + 
  ggtitle(paste("Correlation between the two set of estimates: ", round(
    cor(cases_f_weib_e$S, cases_f_weib_e$S_w), 6))
    )+ 
  
  theme_minimal()

```

